Sparse Function-space Representation \\ of Neural Networks

### 2023-05-14 paper/main.tex
merge

\section{Introduction}
\label{sec:intro}
%
Deep learning has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly how to incorporate uncertainty quantification over predictions which is needed in many applications. Secondly, in the sequential data regime data arrives not in batches

@@

%Current state of affairs 
Recent work on techniques (\eg, \cite{list}) for converting trained neural networks from weight space to function space allow a principled mathematical approach for analyzing the behaviour \cite{add}, performing probabilistic inference \cite{add}, and quantifying uncertainty in neural networks \cite{add}. The methods rely on linearizing the neural network and capturing the first two moments in function space in terms of a mean function and covariance function (kernel)---defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates, and they are a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}. 
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

@@

%
\subsection{Linearization gives rise to a Gaussian process} 
\label{sec:nn2gp}
%
Quantifying uncertainty in a Bayesian framework boils down to finding a posterior of $p(\vw \mid \dataset)$ from, which represents the uncertainty over our parameters based on the prior and data. The posterior structure is related to the loss-landscape around the MAP weights $\vw^*$, which induces a correlation structure on functions produced from the neural network's deterministic mapping. Samples from the weight posterior are perturbed versions of $f_{\vw^*}$ characterizing uncertainty in the learned model in the function space. Although the true posterior is intractable given the non-linearities of the neural network sampling techniques available such as Hamiltonian Monte-Carlo (HMC), The uncertainty plotted \cref{fig:banana} shows the induced function space posterior by sampling through the model. However, in most cases, sampling techniques like HMC are not available given the computational burden.

@@

%A Gaussian procss characterizes a distribution over functions in terms of the first two moments. The key quantity of 
%A Gaussian process is the prior covariance function (or kernel function) $\kappa(\vx,\vx')$, where $\vx, \vx'$ are input vectors in $\R^{\mathrm{D}}$. We aim to find a kernel based on the neural network to build the Gaussian process posterior. 
%A GP is fully characterized by a mean $\mu(\cdot)$ and covariance function $\kappa(\cdot,\cdot)$ (kernel).
%If we had the following function parameterization $g(\vx) = \phi(\vx)^\top \vw$ where $\phi(\vx)$ is a feature map, we can convert from weight space prior to function space prior giving the following kernel $\kappa(\vx, \vx') = \delta^{-1} \phi(\vx) \phi^\T(\vx')$ (see Ch.~XXX in \cite{rasmussen2006gaussian} for details).\todo{this part needs to be rewritten, starts talking about a general function parameterisation and then jumps directly to the neural network} 
%\subsection{Neural Network to GP Prior} 
%\label{sec:nn2gp}
%
A Gaussian process characterizes a distribution over functions in terms of the first two moments, a mean $\mu(\cdot)$ and covariance function $\kappa(\cdot,\cdot)$ (kernel). We aim to find a kernel and mean function based on the neural network to build the Gaussian process posterior.  A linear function in terms of parameters can be converted from the weight space to the function space (see Ch.~XXX in \cite{rasmussen2006gaussian} as follows:
\begin{equation} \label{eq:weight_func}
g_\weights(\mathbf{x}_{i}) = \phi^\top\!(\vx) \, \vw \implies \mu(\vx) = \vzeros \quad \kappa(\mathbf{x}, \mathbf{x}') = \frac{1}{\delta} \phi(\vx)^\top \phi(\vx)
\end{equation}
Given an approximate posterior $q(\vw)$, \citet{immerImprovingPredictionsBayesian2021} proposed to make the following linear approximation
$f^{\textrm{lin.}}_{\vw^*}(\vx; \vw) = f_{\vw^*}(\vx) + \Jac{\weights^*}{\mathbf{x}} \, (\vw - \vw^*)$ of the neural network, where $\Jac{\weights}{\mathbf{x}} \coloneqq \nabla_\weights f(\mathbf{x})^\top \in \R^{C \times P}$ is the Jacobian. The approximation can be used with any approximate inference technique but in the paper they focus on the Laplace-GNN. They find that the linear approximation helps to improve predictions by correcting the approximate posterior when sampled through a non-Gaussian likelihoods.

Furthermore, under the linear approximation they showed you can now convert your weight space prior to the function space by combining with \cref{eq:weight_func} meaning the converged neural network GP prior is:
\begin{equation} \label{eq-laplace-approx-function-space}
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  f^{\textrm{lin.}}_{\vw_*}(\vx; \vw = \vzeros) = f_{\vw^*}(\vx) - \Jac{\weights^*}{\mathbf{x}} \vw^* \quad \text{and} \quad
  \kappa(\mathbf{x}, \mathbf{x}')
  = \frac{1}{\delta} \Jac{\weights^*}{\mathbf{x}}\JacT{\weights^*}{\mathbf{x}'},
\end{equation}
%
%where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}) with Jacobian $\Jac{\weights}{\mathbf{x}} \coloneqq \nabla_\weights f(\mathbf{x})^\top \in \R^{C \times P}$. Alternatively, we could have followed the derivation by \citet{immerImprovingPredictionsBayesian2021}, where they perform a local linearization of the neural network  $f^{\textrm{lin.}}_{\vw^*}(\vx; \vw) = f_{\vw^*}(\vx) + \Jac{\weights^*}{\mathbf{x}} \, (\vw - \vw^*)$. Their mean function differ from ours, \cref{eq-laplace-approx-function-space}, as theirs comes from setting  $\mu(\vx) = f^{\textrm{lin.}}_{\vw_*}(\vx; \vw = \vzeros) = f_{\vw^*}(\vx) - \Jac{\weights^*}{\mathbf{x}} \vw^*$, \ie the difference between the linear model approximation and the neural network function at $\vx$. %which does not seem useful as a prior mean function, 
%On the other hand, we set $\mu(\vx) = \vzeros$, as common in GP literature and derived from a feature map conversion. \todo{derived what?}

@@

\paragraph{Neural network dual variables}
As we are using the MAP solution of the neural network loss we will use the Laplace approximation for the dual variables but other techniques could be used. As was detailed in \cite{wilkinson2023bayes, rasmussen2006gaussian}, the dual variables of the Laplace approximation is the same as removing the posterior averaging from \cref{eq:dual_param} meaning:
\begin{equation}
\label{eq:dual_param_laplace}
\hat{\alpha}_i \coloneqq \nabla_{f_i}\log p(y_i \mid f_i),  \quad \text{and} \quad
\hat{\beta}_i \coloneqq - \nabla^2_{f_i f_i}\log p(y_i \mid f_i).
\end{equation}

@@

\subsection{Sparse dual Gaussian process} 
\label{sec:sparse-dual-gp}
%
Sparse Gaussian processes are a way to reduce the computational complexity of a full GP using a low-rank approximation. Many sparsifying approaches exist (see \cite{quinonero2005unifying}) for an overview. Given that we have our dual variables derived from our neural network predictions and kernel function, we could use any approximation. We decide to follow the sparsifying approach of \citet{titsias2009variational} and also used in the DITC approximaton \cite{quinonero2005unifying}, which is to define the marginal distribution as $q_{\vu}(\vf)  = \int p(\vf  \mid \vu) q(\vu)  d\vu$. Given that we have $p(f_i  \mid \vu)$ determined by our GP prior, the goal is to find a $q(\vu)$. As was shown in \cite{adam2021dual}, the posterior under this model has a particular structure similar to \cref{eq:gp_pred}. In the paper, they spot the dual variables' importance for approximate inference but write them in the natural parameter form as this form is more suitable for optimization through natural gradients. In order to make the link to the dual variables defined in \cref{eq:dual_param} we write them in the sparse GP in the dual variable form.

>>>>

\section{Introduction}
\label{sec:intro}
%
Deep learning has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how to incorporate uncertainty quantification over predictions which is needed in many applications. Secondly, in the sequential data regime data arrives not in batches

@@

%Current state of affairs 
Recent work on techniques (\eg, \cite{list}) for converting trained neural networks from weight space to function space allow a principled mathematical approach for analyzing the behaviour \cite{add}, performing probabilistic inference \cite{add}, and quantifying uncertainty in neural networks \cite{add}. The methods rely on linearizing the neural network and capturing the first two moments in function space in terms of a mean function and covariance function (or kernel) --- defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates, and they are a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}. 
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

@@

%
\subsection{Linearization gives rise to a Gaussian process} 
\label{sec:nn2gp}
%
Quantifying uncertainty in a Bayesian framework boils down to finding a posterior of $p(\vw \mid \dataset)$, which represents the uncertainty over our parameters based on the prior and the data. The posterior structure is related to the loss-landscape around the MAP weights $\vw^*$, which induces a correlation structure on functions produced from the neural network's deterministic mapping. Samples from the weight posterior are perturbed versions of $f_{\vw^*}$ characterizing uncertainty in the learned model in the function space. Although the true posterior is intractable given the non-linearities of the neural network, sampling techniques available, such as Hamiltonian Monte-Carlo (HMC), in most cases, are not available given the computational burden.
 The uncertainty plotted \cref{fig:banana} shows the induced function space posterior by sampling through the model.

@@

%A Gaussian procss characterizes a distribution over functions in terms of the first two moments. The key quantity of 
%A Gaussian process is the prior covariance function (or kernel function) $\kappa(\vx,\vx')$, where $\vx, \vx'$ are input vectors in $\R^{\mathrm{D}}$. We aim to find a kernel based on the neural network to build the Gaussian process posterior. 
%A GP is fully characterized by a mean $\mu(\cdot)$ and covariance function $\kappa(\cdot,\cdot)$ (kernel).
%If we had the following function parameterization $g(\vx) = \phi(\vx)^\top \vw$ where $\phi(\vx)$ is a feature map, we can convert from weight space prior to function space prior giving the following kernel $\kappa(\vx, \vx') = \delta^{-1} \phi(\vx) \phi^\T(\vx')$ (see Ch.~XXX in \cite{rasmussen2006gaussian} for details).\todo{this part needs to be rewritten, starts talking about a general function parameterisation and then jumps directly to the neural network} 
%\subsection{Neural Network to GP Prior} 
%\label{sec:nn2gp}
%
A Gaussian process characterizes a distribution over functions in terms of the first two moments, a mean $\mu(\cdot)$ and covariance function $\kappa(\cdot,\cdot)$ (kernel). We aim to find a kernel and mean function based on the neural network to build the Gaussian process posterior.  A linear function in terms of parameters can be converted from the weight space to the function space (see Ch.~XXX in \cite{rasmussen2006gaussian} as follows:
\begin{equation} \label{eq:weight_func}
g_\weights(\mathbf{x}) = \phi^\top\!(\vx) \, \vw \implies \mu(\vx) = \vzeros \quad \kappa(\mathbf{x}, \mathbf{x}') = \frac{1}{\delta} \phi(\vx)^\top \phi(\vx)
\end{equation}
Given an approximate posterior $q(\vw)$, \citet{immerImprovingPredictionsBayesian2021} proposed to make the following linear approximation
$f^{\textrm{lin.}}_{\vw^*}(\vx; \vw) = f_{\vw^*}(\vx) + \Jac{\weights^*}{\mathbf{x}} \, (\vw - \vw^*)$ of the neural network, where $\Jac{\weights}{\mathbf{x}} \coloneqq \nabla_\weights f(\mathbf{x})^\top \in \R^{C \times P}$ is the Jacobian. The approximation can be used with any approximate inference technique but in the paper they focus on the Laplace-GGN. They find that the linear approximation helps to improve predictions by correcting the approximate posterior when sampled through a non-Gaussian likelihoods.

Furthermore, under the linear approximation they showed one can convert the weight space prior to the function space by combining with \cref{eq:weight_func} meaning the converged neural network GP prior is:
\begin{equation} \label{eq-laplace-approx-function-space}
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  f^{\textrm{lin.}}_{\vw_*}(\vx; \vw = \vzeros) = f_{\vw^*}(\vx) - \Jac{\weights^*}{\mathbf{x}} \vw^* \quad \text{and} \quad
  \kappa(\mathbf{x}, \mathbf{x}')
  = \frac{1}{\delta} \Jac{\weights^*}{\mathbf{x}}\JacT{\weights^*}{\mathbf{x}'},
\end{equation}
%
%where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}) with Jacobian $\Jac{\weights}{\mathbf{x}} \coloneqq \nabla_\weights f(\mathbf{x})^\top \in \R^{C \times P}$. Alternatively, we could have followed the derivation by \citet{immerImprovingPredictionsBayesian2021}, where they perform a local linearization of the neural network  $f^{\textrm{lin.}}_{\vw^*}(\vx; \vw) = f_{\vw^*}(\vx) + \Jac{\weights^*}{\mathbf{x}} \, (\vw - \vw^*)$. Their mean function differ from ours, \cref{eq-laplace-approx-function-space}, as theirs comes from setting  $\mu(\vx) = f^{\textrm{lin.}}_{\vw_*}(\vx; \vw = \vzeros) = f_{\vw^*}(\vx) - \Jac{\weights^*}{\mathbf{x}} \vw^*$, \ie the difference between the linear model approximation and the neural network function at $\vx$. %which does not seem useful as a prior mean function, 
%On the other hand, we set $\mu(\vx) = \vzeros$, as common in GP literature and derived from a feature map conversion. \todo{derived what?}

@@

\paragraph{Neural network dual variables}
As we are using the MAP solution of the neural network loss, we will use the Laplace approximation for the dual variables but other techniques could be used. As detailed in \cite{wilkinson2023bayes, rasmussen2006gaussian}, the dual variables of the Laplace approximation are the same as removing the posterior averaging from \cref{eq:dual_param}, meaning:
\begin{equation}
\label{eq:dual_param_laplace}
\hat{\alpha}_i \coloneqq \nabla_{f_i}\log p(y_i \mid f_i),  \quad \text{and} \quad
\hat{\beta}_i \coloneqq - \nabla^2_{f_i f_i}\log p(y_i \mid f_i).
\end{equation}

@@

\subsection{Sparse dual Gaussian process} 
\label{sec:sparse-dual-gp}
%
Sparse Gaussian processes are a way to reduce the computational complexity of a full GP using a low-rank approximation. Many sparsifying approaches exist, we refer the reader to \cite{quinonero2005unifying} for an overview. Given that we have our dual variables derived from our neural network predictions and kernel function, we could use any approximation. We decide to follow the sparsifying approach of \citet{titsias2009variational} and also used in the DITC approximaton \cite{quinonero2005unifying}, which is to define the marginal distribution as $q_{\vu}(\vf)  = \int p(\vf  \mid \vu) q(\vu)  d\vu$. Given that we have $p(f_i  \mid \vu)$ determined by our GP prior, the goal is to find a $q(\vu)$. As was shown in \cite{adam2021dual}, the posterior under this model has a particular structure similar to \cref{eq:gp_pred}. In the paper, they spot the dual variables' importance for approximate inference but write them in the natural parameter form as this form is more suitable for optimization through natural gradients. In order to make the link to the dual variables defined in \cref{eq:dual_param} we write them in the sparse GP in the dual variable form.
